Multi-stage programming with functors and monads: eliminating abstraction overhead from generic code

### 2007-07-13 SCP2008/paper/metamonads.tex
the rest of my comments

The monad represents fine-scale code generation. We need tools for
larger-scale modularization; we can use any abstraction
mechanisms we want to structure our code generators, as long as none
of those abstractions infiltrate the generated code.

\jacques{This whole section needs to be edited to take into account
the different organization of the material}.

\subsection{Domains}
Reviewing the various aspects outlined in section~\ref{design}, the
simplest parametrization is to make the domain abstract.  As it turns
out, we need the following to exist in our domains: $0$, $1$, $+$,
$*$, (unary and binary) $-$, at least \emph{exact} division,
normalization, and potentially a relative size measure. The simplest
case of such domain abstraction is |param_code1|.  There,
code-generators such as |plus| and |one| were passed as arguments. We
need far more than two parameters, so we have to group them. Instead
of the grouping offered by regular records, we use OCaml
\emph{structures} (i.e., modules) so we can take advantage of
extensibility, type abstraction and constraints, and especially
parameterized structures (\emph{functors}).  We define the type of the
domain, the signature |DOMAIN|, which different domains must satisfy:

@@

A more complex domain is |Zp| of the field of integers in prime
modulus:
\begin{code}
module ZpMake(P:sig val p:int end) = struct
    type v = int
    let kind = Domain_is_Field
    let zero = 0 and one = 1
    let plus x y = (x + y) mod P.p
    let times x y = (x * y) mod P.p
    \dots
    let normalizer = None and better_than = None
    let () = assert (is_prime P.p)
end
\end{code}
This domain is parametrized by an integer number |p|, the
(modulus?). To be more precise, the structure |ZpMake| is
parameterized over another structure of the type described by
signature P; that argument structure has one field, the int value |p|.
Such a parameterized structure (or, a function from structures to
structures) is a \emph{functor}. The result of |ZpMake| is a domain that is a
field with no defined order. Hence
|normalizer| and |better_than| are set to |None|. |Zp| forms
a field only when |p| is prime. Therefore, we must make this check,
in the last line of the above code. That line differs from the other 
bindings in |ZpMake| in that it neither defines a function, such as
|plus|, nor binds a value, such as |zero|. Therefore, the non-value
expression |assert (is_prime P.p)| will be evaluated
when the corresponding module is instantiated, like the following
\begin{code}
module Z19 = ZpMake(struct let p = 19 end)
\end{code}
We will call such an expression an initializing expression.  if we
replace |p = 19| with |p = 9| above, we receive an error, a run-time
error. It is raised however as we instantiate and combine modules that
will eventually make the generator.  Although the error is reported at
run-time rather than compile time as one might have hoped, the error
is raised when \emph{generating the generator} -- well before the
generation of the target code could begin. In our code we make
extensive use of these so-called preflight checks performed as part of
module initialization. The checks seem to offer a good compromise:
they are dynamic and so do not require complicated type systems; on
the other hand, the checks are performed quite early, when buiding
code generators, and so ensure that no code violating the
corresponding \emph{semantic} constraints will be generated. Although
some may frown on the use of module initializing expressions (since
that requires careful attention to sharing and multiple instantiations
of a module), these concerns do not apply in our case: our preflight
tests are all idempotent and maintain no state.

@@

This encapsulation of staging annotations into a code combinators lets
us easily change the output form of our generated LU programs. We may
use single-stage thunks rather than MetaOCaml staged |('a,'v) code|
values, for benchmarking and regression tests purposes. We could use
code combinators producing C or Fortran code.

Our code combinator library includes operations to create, dereference
and mutate reference cells, compare values, manipulate values used
for indexing arrays. The following is the implementation of a sample
combinators, in the case of the abstract code being MetaOCaml code values. 
\begin{code}
let lift x = .< x >.

let cunit = .< () >.
let unitL = fun s k -> k s .< () >.

let liftRef x = .< ref .~x >. 
let liftGet x = .< ! .~x >. 
let liftPair x = (.< fst .~x >., .< snd .~x >.)

@@

module Idx = struct
  let zero = .< 0 >. and one = .< 1 >. and minusone = .< -1 >.
  let succ a = .< .~a + 1 >.
  let pred a = .< .~a - 1 >.
  let less a b = .< .~a < .~b >.
  let uminus a = .< - .~a >.
  let add a b = .< .~a + .~b >.

@@

We now convert the domains of values from teh previous section into
lifted domains, of abstract code values. We define the signature
|DOMAINL| 
\begin{code}
module S(T: sig type ('a, 'b) rep  end) = struct
open T
module type DOMAINL = sig
  include DOMAIN
  type 'a vc = ('a,v) rep
  val zeroL : 'a vc
  val oneL : 'a vc
  val ( +^ ) : 'a vc -> 'a vc -> 'a vc
  val ( *^ ) : 'a vc -> 'a vc -> 'a vc
  val ( -^ ) : 'a vc -> 'a vc -> 'a vc
  val uminusL : 'a vc -> 'a vc
  val divL : 'a vc -> 'a vc -> 'a vc
  val better_thanL : ('a vc -> 'a vc -> ('a,bool) rep) option
  val normalizerL : ('a vc -> 'a vc) option
end 
\end{code}
(part of the larger module |Domains_sig|) which is parameterised by
the type of the abstract code, |('a, 'b) rep|. The line |include DOMAIN|
says that lifted domains include all members of non-lifted domains,
including the initializing expressions with pre-flight checks.
non-lifted

The following is a particular instance of |DOMAINL|, the lifted
version of |IntegerDomain| of the previous section, in the particular
case of abstract code being MetaOCaml code values |('a, 'b) code|:
\begin{code}
module T = Domains_sig.S(struct type ('a, 'b) rep = ('a, 'b) code end)
open T
module IntegerDomainL = struct
    include IntegerDomain
    type 'a vc = ('a,v) code
    let zeroL = .< 0 >.  
    let oneL = .< 1 >. 
    let (+^) x y = .<.~x + .~y>. 
    let ( *^ ) x y = .<.~x * .~y>.
    let ( -^ ) x y = .<.~x - .~y>.
    let uminusL x = .<- .~x>.
    let divL x y = .<.~x / .~y>. 
    let normalizerL = None
    let better_thanL = Some (fun x y -> .<abs .~x > abs .~y >. )
end
\end{code}

@@

The signature |CONTAINER2D| specifies that a container must provide
functions |dim1| and |dim2| to extract the dimensions, functions |getL|
to generate container getters, the cloning
generator |copy| and functions that generate code for row and column
swapping. The inclusion of these functions in the signature of all
containers makes it simpler to optimize the relevant functions
depending on the actual representation of the container while not
burdening the users of containers with efficiency details.

@@

The type of our containers include the lifted domain |Dom| as one of
the component. This is quite convenient since operations on containers
are usually accompanied by the operations on the retrieved and stored
values. Our containers are parametric over a |DOMAINL|, i.e., functors
from a |DOMAINL| module to the actual implementation of a
container. For example, the following functor defines a matrix
container represented as arrays of rows.

@@

Various aspects of our LU generator such as determinant
computation, rank, permutation list may need to keep a state during
the code generation. For example, the initializing section of the LU
generator will invoke the |decl| of the determinant aspect so that the
latter may generate let-binding for reference cells that track the
magnitude and the sign of the determinant. The generator of the
row-swapping code will notify the determinant aspect so that the
latter may insert the code that flips the sign of the tracked
determinant. Finally, the final section of the generator will ask the
determinant aspect to generator code producing the final, signed value
of the determinant. Clearly the determinant aspect needs to maintain
state from one invocation of its methods to another; this state should
include the `names' of mutable variables accumulating the sign and the
magnitude of the determinant.

The simplest method of keeping the state of the generator is by using
mutable variables private to the Deteminant module (aspect). That,
however, makes our aspects stateful. Although we are generating
imperative code, we would like to keep our generators stateless and
pure functional, for ease of comprehension and reasoning. Also, if a
module is stateless, we do not have to worry about sharing, aliasing
and multiple instances. Our main program may include several LU
generators, which may refer to the Determinant aspect -- which may be
present in one shared instance or in several. It is irrelevant for us
which is which. With stateful modules, the issues of aliasing or
separate instantiation are the source of very subtle problems.

@@

This monadic state has to accommodate several pieces of state, for
various aspects. We should be able to add a new aspect, which may need
to keep its own state as part of teh overall monadic state, without
modifying or even recompiling the rest of the code. Thus our monadic
state should be extensible. One way of achieving is via an extensible
record, or OCaml object. Each aspect will have its own field; record
subtyping will assure modularity. Alas, this approach makes it
difficult to create the initial state, to pass to the monad's |runM|
method. We should know the names of all fields and should know the
proper initial value for these fields. This breaks the modularity.

A better approach of property list has been suggested my the MLton
team \cite{mlton-proplist}.  Encoding an object by its dual, list of
polymorphic variants.  Property list represents an extensible
object. The initial object is just the empty list. Alas, the MLton's
approach does not literally apply to us. First of all, due to
the environment classifiers, we need universes parameterized by
classifiers. Also, the property names are generated automatically
in MLton approach, based on generativity of exceptions or reference
cells. That means if some particular aspect happens to be included
twice, the state will be incompatible. We will have to worry about
sharing or the lack of sharing of our aspects. We would rather
prefer our aspects to be stateless and work the same way whether two
instances of the aspect are shared or not. Fortunately, OCaml offers a
way to build open unions using polymorphic variants. Each variant is
identified by a manifest tag, e.g., `Tdet, and may include a
value. The tags are not generative and provide the form of manifest
naming (somewhat similar to symbols of Lisp and Scheme).

Below is the implementation of our open records with manifest
naming, functions |orec_store| to add a new field to the open record
and |orec_find| to obtain the value associated with a particular field
name. Each `field' is characterized by a triple: injection, 
projection functions and the string name. The latter is used for 
printing error messages. For example, for the
determinant tracking aspect, this triple has the form
\begin{code}
  let ip = (fun x -> `TDet x), 
           (function `TDet x -> Some x | _ -> None), 
           "Det"
\end{code}
We combine these functions with monadic
actions to access monadic state and so obtain 
|mo_extend| and |mo_lookup| to store and retrieve one particular value
of the monadic state.
\begin{code}
let rec lookup ((_,prj,_) as ip) = 
   function [] -> raise Not_found
   | (h::t) -> (match prj h with Some x -> x | _ -> lookup ip t)

@@

let mo_lookup ip =
  perform s <-- fetch; ret (orec_find ip s)
\end{code}

Currently, we check at generator-time that one cannot add a slot which
already exists (already the part of the state) nor we can obtain the
value of a non-existing slot. These problems may occur if our genetor
is wrong (calls fin method before decl method of an aspect). It is
possible to make these checks static (TFP2007, state-variable monad?)
state-changing monad...

@@

To track determinant we should be able to generate code
for: defining variables used for tracking (|decl|),
updating the sign or the absolute
value of the determinant, converting the tracking state
to the final determinant value of the type |outdet|. LU of a
floating-point matrix with no determinant tracking uses the
instantiation of |DETERMINANT| where |outdet| is |unit| and all the
functions of that module generate no code. For integer matrices, we
have to track some aspects of the determinant, even if we don't output
it. The determinant tracking aspect is complex because tracking
variables, if any, are to be declared at the beginning of LU; the sign
of the determinant has to be updated on each row or column
permutation; the value of the determinant should be updated per each
pivoting. We use |lstate| to pass the tracking state, e.g., a piece of
code for the value of the type |Dom.v ref|, among
various determinant-tracking functions. The |lstate| is a part of the
overall monadic state.

Other aspects follow a similar outline

\begin{code}
module type DETERMINANT = sig
  type tdet = C.Dom.v ref
  type 'a lstate
  type ('pc,'v) lm = ('pc,'v) cmonad
    constraint 'pc = <state : [> `TDet of 'a lstate ]; classif : 'a; ..>
  type ('pc,'v) om = ('pc,'v) omonad
    constraint 'pc = <state : [> `TDet of 'a lstate ]; classif : 'a; ..>
  val decl : unit -> ('b,unit) lm (* could be unit rather than unit code...*)
  val upd_sign  : unit -> ('b,unit) om
  val zero_sign : unit -> ('b,unit) lm
  val acc       : ('a,C.Dom.v) abstract -> (<classif : 'a; ..>,unit) lm
  val get       : unit -> ('b,tdet) lm
  val set       : ('a,C.Dom.v) abstract -> (<classif : 'a; ..>,unit) lm
  val fin       : unit -> ('b,C.Dom.v) lm
end
\end{code}

We have two instances of |Det|. The first one is used when the user
requested that no determinant tracking is needed, and so no code is
generated. The methods just return without executing any
code-generation action.
\begin{code}
module NoDet =
  struct
  type tdet = C.Dom.v ref
  type 'a lstate = unit
  let decl () = unitL
  let upd_sign () = ret None
  let zero_sign () = unitL
  let acc _ = unitL
  let get () = ret (liftRef C.Dom.zeroL)
  let set _ = unitL
  let fin () = failwith "Determinant is needed but not computed"
  type ('pc,'v) lm = ('pc,'v) cmonad
    constraint 'pc = <state : [> `TDet of 'a lstate ]; classif : 'a; ..>
  type ('pc,'v) om = ('pc,'v) omonad
    constraint 'pc = <state : [> `TDet of 'a lstate ]; classif : 'a; ..>
end
\end{code}

@@

\begin{code2}
module type OUTPUTDEP = sig 
    module PivotRep : PIVOTKIND 
    module Det      : DETERMINANT
end
(* The `keyword' list of all the present internal features *)
module type INTERNAL_FEATURES = sig
  module R      : TrackRank.RANK
  module P      : TRACKPIVOT
  module L      : LOWER
end

module type OUTPUT = functor(OD : OUTPUTDEP) -> sig
  module IF : INTERNAL_FEATURES
  type res
  val make_result : 'a wmatrix -> (\dots,res) monad
end
module OutDetRank(OD : OUTPUTDEP) = struct
  module IF = struct
      module R   = Rank
      module P   = DiscardPivot
      module L   = NoLower end
  type res = C.contr * C.Dom.v * int
  let make_result m = perform
    det  <-- OD.Det.fin ();
    rank <-- IF.R.fin ();
    ret (Tuple.tup3 m.matrix det rank)
  (* Initialization: check the preconditions of instantiation of this struct*)
  let _ = OD.Det.fin ()
  let _ = IF.R.fin ()
end
\end{code2}

The initialization expressions |OD.Det.fin ()| and |IF.R.fin ()| are
the pre-flight tests, to be explained in Section~\ref{something}.

Both instances of |Det| contain the |fin ()| function to generate the
code representing the computed determinant. Only the |NoDet| does not
track any determinant, and so |fin ()| raises an error. The code 
|let _ = OD.Det.fin ()| in |OutDetRank| invokes that |fin| function, which will
produce the monadic code generating action, or raise the error. We do
not run the action at that time -- we only make sure there is an
action to run. This is the preflight check again, to rule out the
semantic error when the user specified that determinant should be computed
and returned and yet specified the |NoDet| aspect which tracks no
determinant at all.

\subsection{Main generation}
We combine all aspects in a record. This record is like a keyword
argument list to the |GenGE| functor below.
\begin{code}
module type FEATURES = sig
  module Det       : DETERMINANT
  module PivotF    : PIVOT
  module PivotRep  : PIVOTKIND
  module Update    : UPDATE
  module Input     : INPUT
  module Output    : OUTPUT
end
\end{code}

\oleg{describe how generation proceeds}
We first instantiate the container |GVC_I| for the vector over an
integer domain. We pass the container to the main functor |GenLA|,
since all the linear algebra aspects are parameterized by the
container and the domain included in it. We open the |G_GVC_I| module
to make all its components easily available. One of thos components is
the GE generrator, |GE| (which we open as well). The module |G_GVC_I|
also inlcudes the generator for GE solvers. The module |GE| includes
the functor |GenGE| that is the generator. We instantiate the
generator by passing the functor the `record' of various aspects (the
order is irrelevant, but all aspects like |Det|, etc. must be
specified). In the code below, we request generation with full pivot,
fraction-free update, operate on non-augmented matrix and return the
|U| factor, determinant and the rank. We run the monad passing the
initial state |[]| and obtain the code in App A.

\begin{code}
module GVC_I = GenericVectorContainer(IntegerDomainL)
module G_GVC_I = GenLA(GVC_I)
open G_GVC_I
open G_GVC_I.GE
module GenIV5 = GenGE(struct 
    module Det = AbstractDet
    module PivotF = FullPivot
    module PivotRep = PermList
    module Update = FractionFreeUpdate
    module Input = InpJustMatrix
    module Output = OutDetRank end)
let instantiate gen =
    .<fun a -> .~(runM (gen .<a>.) []) >.;;
let resIV5 = instantiate GenIV5.gen ;;
let rIV5 = .! resIV5;;

module GAC_F = GenericArrayContainer(FloatDomainL)
module G_GAC_F = GenLA(GAC_F)
open G_GAC_F
open G_GAC_F.GE
module GenFA9 = GenGE(struct 
    module Det = NoDet
    module PivotF = RowPivot
    module PivotRep = PermList
    module Update = DivisionUpdate
    module Input = InpJustMatrix
    module Output = Out_LU_Packed end)
\end{code}

@@

    let zerobelow mat pos = 
        let module IF = O.IF in
        let module U = F.Update(F.Det) in
        let innerbody j bjc = perform
            whenM (Logic.notequalL bjc C.Dom.zeroL ) (perform
                det <-- F.Det.get ();
                optSeqM (Iters.col_iter mat.matrix j (Idx.succ pos.p.colpos) 
               (Idx.pred mat.numcol) C.getL
                      (fun k bjk -> perform
                      brk <-- ret (C.getL mat.matrix pos.p.rowpos k);
                      U.update bjc pos.curval brk bjk 
                          (fun ov -> C.col_head_set mat.matrix j k ov) det) UP )
                      (IF.L.updt mat.matrix j pos.p.colpos C.Dom.zeroL 
                          (* this makes no sense outside a field! *)
                          (C.Dom.divL bjc pos.curval))) in
        perform
              seqM (Iters.row_iter mat.matrix pos.p.colpos
              (Idx.succ pos.p.rowpos)
              (Idx.pred mat.numrow) C.getL innerbody UP) 
                   (U.update_det pos.curval)

>>>>

The monad represents fine-scale code generation. We need tools for
larger-scale modularization; we can use any abstraction
mechanisms we want to structure our code generators, as long as none
of those abstractions infiltrate the generated code. ML module system
turns out most convenient.

\subsection{Domains}

We start by abstracting out domains. We have already seen the simplest
case of domain abstraction in |param_code1|, which took
code-generators such as |plus| and |one| as arguments.  We need far
more than two parameters: our domains should include $0$, $1$, $+$,
$*$, (unary and binary) $-$, at least \emph{exact} division,
normalization, and potentially a relative size measure. We could group
these parameters in tuples or records.  Instead, we use OCaml
\emph{structures} (i.e., modules) so we can take advantage of
extensibility, type abstraction and constraints, and especially
parameterized structures (\emph{functors}).  We define the type of the
domain, the signature |DOMAIN|, which different domains must satisfy:

@@

A more complex domain is |Zp| of the field of integers in prime
modulus:
\begin{code}
module ZpMake(P:sig val p:int end) = struct
    type v = int
    let kind = Domain_is_Field
    let zero = 0 and one = 1
    let plus x y = (x + y) mod P.p
    let times x y = (x * y) mod P.p
    \dots
    let normalizer = None and better_than = None
    let () = assert (is_prime P.p)
end
\end{code}
This domain is parametrized by an integer number |p|. 
To be more precise, the structure |ZpMake| is
parameterized over another structure of the type described by
the signature P; that argument structure has one field, the |int| value |p|.
Such a parameterized structure (or, a function from structures to
structures) is a \emph{functor}. The result of |ZpMake| is a domain that is a
field with no defined order. Hence
|normalizer| and |better_than| are set to |None|. |Zp| forms
a field only when |p| is prime. Therefore, we must make this check,
in the last line of the above code. That line differs from the other 
bindings in |ZpMake| in that it neither defines a function, such as
|plus|, nor binds a value, such as |zero|. Therefore, the non-value
expression |assert (is_prime P.p)| will be evaluated
when the corresponding module is instantiated, as in the following
\begin{code}
module Z19 = ZpMake(struct let p = 19 end)
\end{code}
We call such an expression an initializing expression. If we
replace |p = 19| with |p = 9| above, we receive a run-time
error. It is raised however as we instantiate and combine modules that
will eventually make the generator.  Although the error is reported at
run-time rather than compile time as one might have hoped, the error
is raised when \emph{generating the generator} -- well before the
generation of the target code could begin. In our code we make
extensive use of these so-called preflight checks performed as part of
module initialization. The checks seem to offer a good compromise:
they are dynamic and so do not require complicated type systems; on
the other hand, the checks are run quite early, when buiding
code generators, and so ensure that no code violating the
corresponding \emph{semantic} constraints will be generated. Although
some may frown on the use of module initializing expressions since
that requires careful attention to sharing and multiple instantiations
of a module, these concerns do not apply in our case: our preflight
checks are all idempotent and maintain no state.

@@

This encapsulation of staging annotations into code combinators lets
us easily change the output form of our generated LU programs. We may
use single-stage thunks rather than MetaOCaml staged |('a,'v) code|
values, for benchmarking and regression tests purposes. We could use
code combinators producing C or Fortran code.

Our code combinator library includes operations to create, dereference
and mutate reference cells, compare values, manipulate values used
for indexing arrays. The following is the implementation of sample
combinators, in the case of the abstract code being MetaOCaml code. 
\begin{code}
let lift x = .< x >.

let cunit = .< () >.  and unitL = fun s k -> k s .< () >.

let liftRef x = .< ref .~x >.  and   liftGet x = .< ! .~x >. 
let liftPair x = (.< fst .~x >., .< snd .~x >.)

@@

module Idx = struct
  let zero = .< 0 >. and one = .< 1 >. and minusone = .< -1 >.
  let succ a = .< .~a + 1 >.  and  pred a = .< .~a - 1 >.
  let less a b = .< .~a < .~b >.
  let uminus a = .< - .~a >.  and  add a b = .< .~a + .~b >.

@@

We now convert the domains of values from the previous section into
lifted domains, of abstract code values. We define the signature
|DOMAINL| 
\begin{code}
module S(T: sig type ('a, 'b) rep  end) = struct
open T
module type DOMAINL = sig
  include DOMAIN
  type 'a vc = ('a,v) rep
  val zeroL : 'a vc
  val oneL : 'a vc
  val ( +^ ) : 'a vc -> 'a vc -> 'a vc
  val ( *^ ) : 'a vc -> 'a vc -> 'a vc
  val ( -^ ) : 'a vc -> 'a vc -> 'a vc
  val uminusL : 'a vc -> 'a vc
  val divL : 'a vc -> 'a vc -> 'a vc
  val better_thanL : ('a vc -> 'a vc -> ('a,bool) rep) option
  val normalizerL : ('a vc -> 'a vc) option
end 
\end{code}
(part of the larger module |Domains_sig|) which is parameterised by
the type of the abstract code, |('a, 'b) rep|. The line |include DOMAIN|
says that lifted domains include all members of non-lifted domains,
including the initializing expressions with pre-flight checks.
The following is a particular instance of |DOMAINL|, the lifted
version of |IntegerDomain| of the previous section, in the particular
case of abstract code being MetaOCaml code values |('a, 'b) code|:
\begin{code}
module T = Domains_sig.S(struct type ('a, 'b) rep = ('a, 'b) code end)
open T
module IntegerDomainL = struct
    include IntegerDomain
    type 'a vc = ('a,v) code
    let zeroL = .< 0 >.  
    let oneL = .< 1 >. 
    let (+^) x y = .<.~x + .~y>. 
    let ( *^ ) x y = .<.~x * .~y>.
    let ( -^ ) x y = .<.~x - .~y>.
    let uminusL x = .<- .~x>.
    let divL x y = .<.~x / .~y>. 
    let normalizerL = None
    let better_thanL = Some (fun x y -> .<abs .~x > abs .~y >. )
end
\end{code}

@@

The signature |CONTAINER2D| below specifies that a container must provide
functions |dim1| and |dim2| to extract the dimensions, functions |getL|
to generate container getters, the cloning
generator |copy| and functions that generate code for row and column
swapping. The inclusion of these functions in the signature of all
containers makes it simpler to optimize the relevant functions
depending on the actual representation of the container while not
burdening the users of containers with efficiency details.

@@

The type of our containers include the lifted domain |Dom| as one of
the components. This is quite convenient since operations on containers
are usually accompanied by the operations on the retrieved and stored
values. Our containers are parametric over a |DOMAINL|, i.e., functors
from a |DOMAINL| module to the actual implementation of a
container. For example, the following functor defines a matrix
container represented as arrays of rows.

@@

Various aspects of our LU generator such as determinant, rank, 
and permutation traking may need to keep a state during
the code generation. For example, the initializing section of the LU
generator will invoke the function |decl| of the determinant aspect so that the
latter may generate let-binding for reference cells that track the
magnitude and the sign of the determinant. The generator of the
row-swapping code will notify the determinant aspect so that the
latter may insert the code that flips the sign of the tracked
determinant. The final section of the LU generator will ask the
determinant aspect to generate code producing the final, signed value
of the determinant. Clearly the determinant aspect needs to maintain
state from one invocation of its functions to another; this state should
include the `names' of mutable variables accumulating the sign and the
magnitude of the determinant.

The simplest method of keeping the state of the generator is by using
mutable variables private to the Deteminant module (aspect). That,
however, makes our aspects stateful. Although we are generating
imperative code, we would like to keep our generators stateless and
pure functional, for ease of comprehension and reasoning.  Our main
program may include several LU generators referring to one Determinant
aspect -- which may be present in one shared instance or in
several. That is of no concern if the Determinant module is
stateless. With stateful modules, the issues of aliasing or separate
instantiation are the source of very subtle problems.

@@

This monadic state has to accommodate several pieces of state, for
various aspects. We should be able to add a new aspect -- which may need
to keep its own state as part of the overall monadic state -- without
modifying or even recompiling the rest of the code. Thus our monadic
state should be extensible. We could use an extensible
record: an OCaml object. Each aspect will have its own field; record
subtyping will assure modularity. Alas, this approach makes it
difficult to create the initial state, to pass to the monad's |runM|
method. We should know the names of all fields and should know the
proper initial value for these fields, which breaks modularity.

A better approach of a property list has been suggested by the MLton
team \cite{mlton-proplist}. A property list, too, represents an
extensible object -- by its dual, a list of polymorphic variants.  The
initial object is just the empty list. MLton's approach does not
literally apply to us however. Environment classifiers of MetaOCaml
require type universes parameterized by classifiers. The MLton
approach uses generativity of exceptions or reference cells to
generate property names, i.e., fields of an extensible record. That
would make our aspects stateful modules, which we try to
avoid. Fortunately, OCaml offers a way to build open unions with
polymorphic variants. Each variant is identified by a manifest tag,
e.g., |`Tdet|, and may include a value. The tags are not generative and
provide a form of manifest naming, similar to symbols of
Lisp and Scheme.

Below is the implementation of our open records with manifest
naming: functions |orec_store| to add a new field to the open record
and |orec_find| to obtain the value associated with a particular field
name. Each `field' is characterized by a triple: injection, 
projection functions and the string name. The latter is used for 
printing error messages. For example, for the
determinant tracking aspect, this triple has the form
\begin{code}
let ip = 
  (fun x -> `TDet x), (function `TDet x -> Some x | _ -> None), "Det"
\end{code}
We combine these functions with monadic
actions to access monadic state and so obtain 
|mo_extend| and |mo_lookup| to store and retrieve one component 
of the monadic state.
\begin{code}
let rec lookup ((_,prj,_) as ip) = 
   function [] -> raise Not_found
   | (h::t) -> (match prj h with Some x -> x | _ -> lookup ip t)

@@

let mo_lookup ip = perform s <-- fetch; ret (orec_find ip s)
\end{code}
%
Currently, we check at generator-time that one should not add an
already existing field to an open record, nor should one attempt to
look up a field that does not exist. It is possible to make these
checks static.

@@

To track determinant we should be able to generate code
for: defining variables used for tracking (|decl|),
updating the sign or the absolute
value of the determinant, converting the tracking state
to the final determinant value of the type |outdet|. LU of a
floating-point matrix with no determinant tracking uses the
instantiation of |DETERMINANT| where |outdet| is |unit| and all the
functions of that module generate no code. For integer matrices, we
have to track some aspects of the determinant, even if we don't output
it. The determinant tracking aspect is complex because tracking
variables, if any, are to be declared at the beginning of LU; the sign
of the determinant has to be updated on each row or column
permutation; the value of the determinant should be updated per each
pivoting. We use |lstate| to pass the tracking state, e.g., a piece of
code for the value of the type |Dom.v ref|, among
various determinant-tracking functions. The |lstate| is a part of the
overall monadic state. 
Other aspects follow a similar outline

\begin{code}
module type DETERMINANT = sig
  type tdet = C.Dom.v ref
  type 'a lstate
  type ('pc,'v) lm = ('pc,'v) cmonad
    constraint 'pc = <state : [> `TDet of 'a lstate ]; classif : 'a; ..>
  type ('pc,'v) om = ('pc,'v) omonad
    constraint 'pc = <state : [> `TDet of 'a lstate ]; classif : 'a; ..>
  val decl : unit -> ('b,unit) lm
  val upd_sign  : unit -> ('b,unit) om
  val zero_sign : unit -> ('b,unit) lm
  val acc       : ('a,C.Dom.v) abstract -> (<classif : 'a; ..>,unit) lm
  val get       : unit -> ('b,tdet) lm
  val set       : ('a,C.Dom.v) abstract -> (<classif : 'a; ..>,unit) lm
  val fin       : unit -> ('b,C.Dom.v) lm
end
\end{code}

We have two instances of |Det|. The first one is used when the user
requested that no determinant tracking is needed, and so no code is
generated. The methods just return without executing any
code-generation action.
\begin{code}
module NoDet =
  struct
  type tdet = C.Dom.v ref
  type 'a lstate = unit
  let decl () = unitL
  let upd_sign () = ret None
  let zero_sign () = unitL
  let acc _ = unitL
  let get () = ret (liftRef C.Dom.zeroL)
  let set _ = unitL
  let fin () = failwith "Determinant is needed but not computed"
end
\end{code}

@@

\begin{code2}
module type OUTPUTDEP = sig 
    module PivotRep : PIVOTKIND 
    module Det      : DETERMINANT
end
module type INTERNAL_FEATURES = sig
  module R      : TrackRank.RANK
  module P      : TRACKPIVOT
  module L      : LOWER
end
module type OUTPUT = functor(OD : OUTPUTDEP) -> sig
  module IF : INTERNAL_FEATURES
  type res
  val make_result : 'a wmatrix -> (\dots,res) monad
end

module OutDetRank(OD : OUTPUTDEP) = struct
  module IF = struct
      module R   = Rank
      module P   = DiscardPivot
      module L   = NoLower end
  type res = C.contr * C.Dom.v * int
  let make_result m = perform
    det  <-- OD.Det.fin ();
    rank <-- IF.R.fin ();
    ret (Tuple.tup3 m.matrix det rank)
  let _ = OD.Det.fin ()
  let _ = IF.R.fin ()
end
\end{code2}

The initialization expressions |OD.Det.fin ()| and |IF.R.fin ()| are
the pre-flight checks. As we saw in the previous section, 
both instances of |Det| contain the |fin ()| function to generate the
code representing the computed determinant. Only the |NoDet| does not
track any determinant, and so |fin ()| raises an error. The code 
|let _ = OD.Det.fin ()| in |OutDetRank| invokes that |fin| function, which will
produce the monadic code generating action, or raise the error. We do
not run the action at that time -- we only make sure there is an
action to run. This is the preflight check again, to rule out the
semantic error when the user specified that determinant should be computed
and returned and yet specified the |NoDet| aspect which tracks no
determinant at all.

\subsection{Main generation}
We combine all aspects in a record, which serevs as a keyword
argument list to the |GenGE| functor below.
\begin{code}
module type FEATURES = sig
  module Det       : DETERMINANT
  module PivotF    : PIVOT
  module PivotRep  : PIVOTKIND
  module Update    : UPDATE
  module Input     : INPUT
  module Output    : OUTPUT
end
\end{code}

\begin{code}
module GVC_I = GenericVectorContainer(IntegerDomainL)
module G_GVC_I = GenLA(GVC_I)
open G_GVC_I
open G_GVC_I.GE
module GenIV5 = GenGE(struct 
    module Det = AbstractDet
    module PivotF = FullPivot
    module PivotRep = PermList
    module Update = FractionFreeUpdate
    module Input = InpJustMatrix
    module Output = OutDetRank end)
let instantiate gen =
    .<fun a -> .~(runM (gen .<a>.) []) >.;;
let resIV5 = instantiate GenIV5.gen ;;
\end{code}
%
We first instantiate the container |GVC_I| for the vector over an
integer domain. We pass the container to the main functor |GenLA|,
since all the linear algebra aspects are parameterized by the
container and the domain included in it. We open the |G_GVC_I| module
to make all its components easily available. One of those components
is the GE generator, |GE| (which we open as well). Besides |GE|, the
module |G_GVC_I| includes the generator for GE solvers.  The main
generator functor is |GenGE|, shown later.  We instantiate the
generator by passing the functor the `record' of various aspects; the
order is irrelevant, but all aspects like |Det|, etc. must be
specified. In the code above, we request GE generation with full
pivot, fraction-free update, operate on non-augmented matrix and
return the |U| factor, determinant and the rank. We run the monad
passing the initial state |[]| and obtain the code, which we can see
by printing |resIV5|. The code can then be ``compiled'' as |!. resIV5|
or with off-shoring. The code for |resIV5| (Appendix A) shows full
pivoting, determinant and rank tracking. The code for all these
aspects is fully inlined; no extra functions are invoked and no tests
other than those needed by the LU algorithm itself are performed. The
resulting function returns a triple |int array * int * int| of the
U-factor, determinant and the rank.

The following is another instantiation of the GE generator, with a
different set of aspects.
\begin{code}
module GAC_F = GenericArrayContainer(FloatDomainL)
module G_GAC_F = GenLA(GAC_F)
open G_GAC_F
open G_GAC_F.GE
module GenFA9 = GenGE(struct 
    module Det = NoDet
    module PivotF = RowPivot
    module PivotRep = PermList
    module Update = DivisionUpdate
    module Input = InpJustMatrix
    module Output = Out_LU_Packed end)
\end{code}
The code generated by |GenFA9| (Appendix B) shows no traces of
determinant tracking whatsoever: no declaration of spurious variables,
no extra tests, etc. The code appears as if the determinant tracking
aspect did not exist at all. The generated code for the above and
other instantiations of |Gen| can be examined at
\cite{metamonadsURL}. The website also contains benchmark code and
timing comparisons.

@@

    let zerobelow mat pos = 
        let module IF = O.IF in
        let module U = F.Update(F.Det) in
        let innerbody j bjc = perform
            whenM (Logic.notequalL bjc C.Dom.zeroL ) (perform
                det <-- F.Det.get ();
                optSeqM (Iters.col_iter mat.matrix j (Idx.succ pos.p.colpos) 
               (Idx.pred mat.numcol) C.getL
                      (fun k bjk -> perform
                      brk <-- ret (C.getL mat.matrix pos.p.rowpos k);
                      U.update bjc pos.curval brk bjk 
                          fun ov -> C.col_head_set mat.matrix j k ov) det) UP )
                      (IF.L.updt mat.matrix j pos.p.colpos C.Dom.zeroL 
                          (* this makes no sense outside a field! *)
                          (C.Dom.divL bjc pos.curval))) in
        perform
              seqM (Iters.row_iter mat.matrix pos.p.colpos
              (Idx.succ pos.p.rowpos)
              (Idx.pred mat.numrow) C.getL innerbody UP) 
                   (U.update_det pos.curval)
